Prediction of Protein-Protein Interactions on the Human and Rice Interactomes

### 2021-01-21 nicolas/PPIPL3/bmc_template/abstract.tex
Recent structure update

\parttitle{Background}
Recent research on network-based prediction of protein-protein
interactions (PPIs) reveals that two proteins are more likely to interact,
the higher the number of paths of length 3 between them exist. This paper
extends previous work on mapping binary interactions between proteins 
by learning features (i.e. embeddings) of the PPI network. In particular, 
this paper reports on an application of gradient boosted decision tree
model (XGBoost) using handcrafted features  such as normalized measures 
and algorithmic embeddings that generate a low-dimensional representation
of the proteins and their interactions (node2vec).

\parttitle{Results}
The main result shows that while the measure remains an important feature
for predicting interactions, an improvement in performance is achieved when
embedding features are considered. The proposed approach is validated for
the human and rice interactomes. For both cases, the combination of both
types of features yield higher performance in terms of the Area Under the
Curve metric (i.e. AUC).

\parttitle{Conclusions}
As witnessed by both human and rice, when information from 
handcrafted features based on neighborhood is enhanced with vector
representations from random walks, the prediction power of the model
improves. Moreover, a supervised learning model has been trained to predict
unknown interactions based on such information. Finally, the developed
framework could also be applied to interactomes of other organisms for which
PPI networks have recently become available.

>>>>

\parttitle{Background}
Recent research on predicting protein-protein
interactions (PPIs) indicates that two proteins are more likely to interact if 
there exists a higher number of paths of length 3 between them. This paper
extends network-based approaches to map binary interactions between proteins 
by combining hand-crafted features such as the number of paths between proteins 
and learned features (i.e. node embeddings) of the PPI network. In particular, 
the proposed approach introduces a gradient boosted decision tree model 
(XGBoost) usese node2vec to infer embeddings, that is, a low-dimensional representation
of the proteins and their interactions.

\parttitle{Results}
The main result shows that while the normalized measure remains an important feature
for predicting interactions, embeddings may improve performance of predictions by up 
to 15\% in terms of the Area Under the Curve metric (i.e. AUC). The approach is validated for 
the human and rice interactomes.

\parttitle{Conclusions}
As shown by the two case studies in human and rice, combining information from 
handcrafted features with embeddings enhances the prediction power of the model. 
Improvements of supervised learning models (such as XGBoost) depend on being able 
to extract additional information about the local neighborhood of nodes that 
complement the measure based on the number of paths. Finally, the proposed 
approach could also be applied to interactomes of a number of other organisms for which
PPI networks have recently become available.

### 2021-01-21 nicolas/PPIPL3/bmc_template/framework.tex
Recent structure update

\subsection*{Framework conception and configuration}

The proposed framework for prediction of protein-protein interactions

\subsection*{Framework evaluation on the human interactome}

As the first step in the evaluation of A2, A3 and L3 methods for link
prediction, the full calculation and ranking of all predictions was
carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
the x axis represents a rank position $k$ and the y axis displays
the precision for the top $k$ predictions of each method, when assuming
interactome \emph{HI-III} as the validation set.

@@

However, it is of our interest to assess if Machine Learning can boost
the overall performance on the prediction, so the proposed procedure
was adapted to the human interactome data: Instead of having to remove
a fraction of edges from the network in order to predict and validate,
the prediction network and the validation networks are given beforehand.
The rest of the procedure is carried out as mentioned in Section 2.
Table \ref{T1} presents the results for the combinations of Node2Vec
with each feature.

>>>>

\subsection*{Framework Conception and Configuration}

As mentioned before, the functionality of a protein strongly depends on its 3D structure, 
so that changes in its conformation (mutations) can affect how a protein interacts with its 
surroundings \cite{Sikosek2014Prot}. In this aspect, $A_G$ gives intrinsic information 
about complementarity between pairs of proteins and hence paths on the protein network are 
relevant for the prediction task.

The proposed framework for prediction of protein-protein interactions consists of the
combination of the path counting metrics along with the vector representations of the
edges of the network (node2vec embeddings). Namely, the path counting metrics are either 
the counting of paths of length 2 (matrix $A_G^2$), the counting of paths of length 3 
(matrix $A_G^3$) or the degree-normalized counting of paths of length 3 (matrix $L_3$). 
The information retrieved from these matrices will be addressed from now on as A2, A3 
and L3, respectively.

Figure \ref{fig:framework} shows the proposed framework for predicting protein-protein interactions. As the 
framework combines two sources of information, it consists of three processes: (a) Metrics 
calculation and initial predictions, (b) Node and edge embeddings and (c) Machine Learning 
prediction.

\begin{figure}[h]
\caption{\label{fig:framework}Framework scheme}
	\includegraphics[width=\textwidth ]{figures/framework.eps}
\end{figure}

The input for the workflow consists of a representation of the connectivity of the network.
Online databases of protein-protein interactions usually report this information as an 
adjacency list with some additional attributes about the quality of this interaction. Namely, 
this data is represented as a list of pairs of identifiers and zero or more attributes for 
each pair. The framework only needs to know whether an interaction exists and therefore this 
additional information needs to be discarded in advance. In order to calculate optimally the
subsequent metrics, the framework creates the adjacency matrix $A_G$ out of the given list. 
Note that as no attributes from nodes or edges are added, the framework predicts based only 
on the intrinsic properties of the network related to connectivity.

\subsubsection*{A. Metrics Calculation}
The goal of this stage is to build matrices $A2$, $A3$ and $L3$ which quantify, 
the number of paths of length 2, of length 3 and the degree-normalized count of 
paths of length 3, respectively. For this purpose, two calculations are done: 
counting the degree of each node ($k_i \in \mathbb{Z}$) and raising $A_G$ to the 
second to obtain $A2$. Finally, $A3$ and $L3$  are calculated simultaneously. 
Entries $L3_{i,j}$ of $L3$ can be calculated as

\begin{equation}
  L3_{i,j}=\sum_{p,q\in V}\frac{{A_G}_{i,p}\cdot {A_G}_{p,q}\cdot {A_G}_{q,j}}{\sqrt{k_{p}\cdot k_{q}}}
\end{equation}

\subsubsection*{B. node2vec Embeddings}
The purpose of this step is to retrieve a vector representation of the network 
structure and thus be able to feed the Machine Learning model and improve its 
predictions. This is done using a state-of-the-art algorithm called node2vec, 
which is based on the word2vec technique to learn learn word associations from
a corpus of text. Instead of sentences, node2vec uses 
random walks from each node in order to construct a skip-gram model and a neural 
network to learn features (embeddings) from the nodes\cite{Grover_2016}. These 
features consist of a $d$-dimensional vector representation of each node.
Since the framework aims to predict edges, these node embeddings are used as 
input for the Hadamard operator to calculate edge embeddings. Namely, the edge 
embedding of edge $c_i,j$ is calculated as the element-wise product of each 
component of the node embeddings for nodes $i$ and $j$. $d$ is chosen as 16.

\subsubsection*{C. Machine Learning Prediction}
Finally, this stage aims at gathering information from the previous steps to 
enrich the prediction of a Machine Learning (ML) model and improve its results.

The selected implementation for this prediction problem is XGBoost, which 
uses gradient boosted decision trees to learn whether an interaction is feasible. 
The first step in this process is to combine the two sources of information into 
Dataframes for each of the metrics: for each edge, the information from one of the 
metrics (A2, A3 or L3) is appended to the vector representation of $d$ dimensions 
from node2vec, resulting in 3 different tables. Dataframes including only one 
source of information (each metric or node2vec data) are also generated and 
fed to XGBoost for comparison.

For each metric, the initial prediction in step A is used to train the supervised 
ML model

\subsection*{Framework Evaluation on the Human Interactome}

As the first step in the evaluation of A2, A3 and L3 metrics to assess link
prediction, the full calculation and ranking of all predictions was
carried out. The results for the analyzed human interactomes (\emph{HI-II-14}
and \emph{HI-TESTED}) is shown in Figure \ref{fig:HI1}, in which
the x axis represents a rank position $k$ and the y axis displays
the precision for the top $k$ predictions of each method, when assuming
interactome \emph{HI-III} as the validation set.

@@

However, the interest of this study is to assess if Machine Learning can boost
the overall performance on the prediction, so the proposed procedure
is adapted to the human interactome data: Instead of having to remove
a fraction of edges from the network in order to predict and validate,
the prediction network and the validation networks are given beforehand.
The rest of the procedure is carried out as mentioned above.
Table \ref{T1} presents the results for the combinations of Node2Vec
with each feature.

### 2021-01-21 nicolas/PPIPL3/bmc_template/introduction.tex
Recent structure update

Proteins are key actors of biological processes inside cells. Rather
than carrying out tasks as single agents, they are part of dynamic
networks of protein-protein interactions (PPI) \cite{Lin2017}. Such
networks underlie a variety of interdependent mechanisms, including
signal transduction, homeostasis control and stress responses. Furthermore,
PPI networks play an important role in physiological and developmental
processes such as protein phosphorylation, transcriptional co-factor
recruitment and transporter activation \cite{Zhang2010PPI}.

A common way to create PPI networks (or validate particular protein-protein
interactions) is the \emph{Yeast-Two-Hybrid} technique (also known
as \emph{two-hybrid screening} or \emph{Y2H}). Figure \ref{Y2H}A
illustrates the biological basis of Y2H: the expression of a specific
reporter gene is activated by the binding of a DNA-binding Domain
(DB) and an Activation Domain (AD) of a Transcription Factor, which
in turn binds to an Upstream Activation Sequence (UAS). To evaluate
an interaction between two proteins, the Y2H approach fuses one protein
to the DB domain (known as \emph{bait}) and the other to the
AD (known as \emph{prey}). If the proteins interact, the reporter
gene expression is activated by the AD (Fig. \ref{Y2H}B). Otherwise,
if proteins fail to interact, the reporter gene is not expressed (Fig.
\ref{Y2H}C).

\begin{figure}[h]
\caption{\label{Y2H}The Yeast-2-Hybrid technique offers an experimental approach
for constructing PPI networks.}
\centering
	\includegraphics[width=0.7\columnwidth]{../Y2H}
\end{figure}

Based on the outcome of Y2H experiments, numerous networks
of interactions between proteins have been constructed \cite{Weimann2013Y2H,Rajagopala2015Y2H,Shokri2019TFY2H}.
However, identifying protein interactions through Y2H experiments is costly in terms of time
and resources\cite{Laraia2015PPI,Macalino2018PPI}. For an organism with 2000
unique proteins, for example, a detailed PPI network requires the experimental
validation of around 2 million interactions. The number of missing
interactions for pairs of proteins on many organisms require the
use of computational tools for predicting such interactions.

Different methods for predicting interactions in PPI networks have
been proposed \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}. A 
common approach for predicting PPIs is based the triadic closure principle
(TCP). TCP states that the higher the number of common neighboring
nodes between two nodes, the higher the probability that they interact
\cite{Goldberg2003SmallWorld}. TCP can be addressed mathematically
by counting the number of shared neighbors of a pair of nodes, also
known as the Common Neighbors algorithm. By raising the adjacency
matrix (A) of the PPI network to the second power (A\texttwosuperior ),
the provides an insight into which proteins may interact with each other. However, previous studies
show that this mentioned approach alone usually fails because it does not
consider the structural and chemical properties of the proteins \cite{Cannistraci2013Networks,Kovacs2019}.

Kovacs et al (2019) introduce a network-based approach which predicts
an interaction between two proteins based on the number of paths
of length 3. The number of paths of length 3 is computed by raising
the adjacency matrix of the network to the third power (A\textthreesuperior).
The degree-normalized measure, denoted as L3,  mitigates bias caused by
intermediate hubs within the paths of length 3. The measure L3 enables
the approach in Kovacs et al (2019) to outperform previous methods for predicting
binary protein interactions for various organisms, including yeast (\emph{S. cerevisiae}),
Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans
\cite{Kovacs2019}.

This paper extends previous analyses by using handcrafted features (A2,
A3, L3) as well as a low-dimensional representation of nodes 
(\texttt{Node2Vec})\cite{Grover_2016} for PPI networks in the human and rice 
interactomes. To evaluate the proposed approach, prediction of interactions is carried on 
with a state-of-the-art Machine Learning algorithm called \texttt{XGBoost}\cite{2016ChenXGB},
which uses a gradient boosted decision tree model to weigh the features.
The main results show that adding the handcrafted features derived from the
network connectivity to the Machine Learning model improves the prediction
power of the models. This property can be assessed when looking at the their
importance in the model.  The main contributions of this paper are i) a
general framework for link prediction in non-directed networks and ii) two
applications of this framework to biological networks with a structural
insight.

\paragraph*{Document structure:} This paper is organized as follows. \emph{Preliminaries} 
gives some foundations on protein-protein interactions, graph theory 
and machine learning. \emph{Prediction Framework} describes
the methodological steps, key milestones and relevant results in the preparation 
of the networks, model parameters and experimental configurations for the human
interactome.
\emph{Case Study} gives a view at the main aspects when exploring the rice
interactome with the aforementioned framework, including experimental setup and results.
\emph{Related Work} compares the framework with some results from the literature, 
exposing some similarities and differences.
\emph{Conclusion} presents the main conclusions of this study. Finally, the \emph{Appendix}
presents supplementary information and figures which complement the
results described in this paper.

>>>>

Proteins are key actors of biological processes inside cells. Rather
than carrying out tasks as single agents, they are part of dynamic
networks of protein-protein interactions (PPIs) \cite{Lin2017}. Such
networks underlie a variety of interdependent mechanisms, including
signal transduction, homeostasis control and stress responses, and 
play an important role in physiological and developmental
processes such as protein phosphorylation, transcriptional co-factor
recruitment and transporter activation \cite{Zhang2010PPI}.

\emph{Yeast-Two-Hybrid} technique (also known as 
\emph{two-hybrid screening} or \emph{Y2H}) is a commonly used 
technique to generate PPI networks. Figure \ref{Y2H}A
illustrates the biological basis of Y2H. The expression of a specific
reporter gene is activated by binding a DNA-binding Domain
(DB) and an Activation Domain (AD) of a Transcription Factor, which
in turn binds to an Upstream Activation Sequence (UAS). To evaluate
an interaction between two proteins, the Y2H approach fuses one protein
(known as \emph{bait}) to the DB domain and the other (known as 
\emph{prey}) to the AD. If the proteins interact, the reporter
gene expression is activated by the AD (Fig. \ref{Y2H}B). Otherwise,
if proteins fail to interact, the reporter gene is not expressed (Fig.
\ref{Y2H}C).

\begin{figure}[h]
\caption{\label{Y2H}Yeast-2-Hybrid offers an experimental approach
for constructing PPI networks.}
\centering
	\includegraphics[width=0.7\columnwidth]{../Y2H}
\end{figure}

Based on the outcome of Y2H experiments, numerous networks
of interactions between proteins have been constructed \cite{Weimann2013Y2H,Rajagopala2015Y2H,Shokri2019TFY2H}.
However, identifying protein interactions through Y2H experiments is 
costly both in terms of time
and resources \cite{Laraia2015PPI,Macalino2018PPI}. For example, 
constructing a PPI network for an organism with 2000 unique 
proteins requires the above validation procedure to be repeated 
around 2 million times. Computational tools for predicting such 
interactions offer a cpst-effective alternative, especially for 
organisms that produce a large number of proteins.

Different methods for predicting interactions based on PPI networks have
been proposed \cite{Chang2016PPI,Chen2019PPI,Kotlyar2015PPI}. A first 
approach for predicting PPIs is based the triadic closure principle
(TCP), which states that the higher the number of common neighboring
nodes between two nodes, the higher the probability that they interact
\cite{Goldberg2003SmallWorld}. TCP is measured by counting the number 
of shared neighbors of a pair of nodes (denoted as A2), namely by raising the adjacency
matrix of the PPI network to the second power ($A_G^2$). 
However, previous studies show that TCP often falls short as it does 
not consider the structural and chemical properties of the proteins
\cite{Cannistraci2013Networks}, which are related to 
complementarity in charges and arrangement of atoms between pairs of proteins.

The work in \cite{Kovacs2019} introduces another network-based approach which predicts
an interaction between two proteins based on the number of paths
of length 3, in contrast to the number of paths of length 2, denoted as A2. 
The number of paths of length 3, denoted as A3, is computed by raising
the adjacency matrix of the network to the third power. Furthermore, the 
degree-normalized measure of A3, denoted as L3,  mitigates bias caused by
intermediate hubs within the paths of length 3. Using the measure of L3, 
the approach in \cite{Kovacs2019} outperforms previous methods for predicting
binary protein interactions for various organisms, including yeast (\emph{S. cerevisiae}),
Arabidopsis (\emph{A. thaliana}), worm (\emph{C. elegans}), fly (\emph{D. melanogaster}),
fission yeast (\emph{S. pombe}), mouse (\emph{M. musculus}) and humans.

This paper extends previous analyses by using handcrafted features (A2,
A3, L3) as well as learned representations of the proteins and their 
neighborhoods within the PPI networks in human and rice 
interactomes (\texttt{node2vec}) \cite{Grover_2016}. To evaluate the 
proposed approach, predictions of interactions are carried out 
using a supervised Machine Learning algorithm called \texttt{XGBoost} 
\cite{2016ChenXGB}. XGBoost is particularly useful since its gradient 
boosted decision tree model weighs the different features and 
establishes the importance of each feature. The main results of this paper 
show that adding the handcrafted features derived from the
network connectivity to the Machine Learning model improves the prediction
power of the models. This property can be assessed when looking at the their
importance in the model. The main contributions of this paper are twofold. 
First, a general framework for link prediction in non-directed networks is 
introduced. Second, two applications of this framework to biological networks 
with a structural insight are presented.

\paragraph*{Document structure:} This paper is organized as follows. The 
section \emph{Preliminaries} gives some foundations on protein-protein 
interactions, graph theory and machine learning. The section 
\emph{Prediction Framework} describes the methodological steps, key 
milestones and relevant results in the preparation of the networks, 
model parameters and experimental configurations for the human
interactome. The section \emph{Case Study} gives a view at the main 
aspects when exploring the rice interactome with the aforementioned 
framework, including experimental setup and results. The section 
\emph{Related Work} compares the framework with some results from the literature, 
exposing some similarities and differences. The final section presents the 
main conclusions of this study. Supplementary information and figures which 
complement the results described in this paper  are presented in the 
\emph{Appendix}.
